### 2018-10-07 paper/abstract.tex
checked fixmes and new material

In this paper, two new contributions for model-based testing with
Communicating Sequential Processes (CSP) are presented. For a given CSP
process representing the reference model, test suites checking the
conformance relations trace refinement and failures refinement are constructed, and
their finiteness and completeness (that is, capability to uncover conformity
violations) is proven. While complete test suites for CSP processes have been
previously investigated by several authors, a sufficient condition for their
finiteness is presented here for the first time. 
%Moreover, an optimal
%technique for testing the admissibility of an implementation's
%nondeterministic behaviour is described. 
It is shown that the test suites are optimal with respect to two aspects:  (a)  the maximal length of test traces cannot be reduced, and (b) the nondeterministic behaviour cannot be tested with smaller or fewer sets of events, without losing the test suite's completeness property. \fixme{alcc: Is the technique
applying the test suites?}
\fxnote{jp: this sentence needed to be re-phrased -- the new version should be better understandable}

>>>>

In this paper, two new contributions for model-based testing with
Communicating Sequential Processes (CSP) are presented. For a given CSP
process representing the reference model, test suites checking the
conformance relations trace refinement and failures refinement are constructed, and
their finiteness and completeness (that is, capability to uncover conformity
violations) is proven. While complete test suites for CSP processes have been
previously investigated by several authors, a sufficient condition for their
finiteness is presented here for the first time.
%Moreover, an optimal
%technique for testing the admissibility of an implementation's
%nondeterministic behaviour is described.
It is shown that the test suites are optimal with respect to two
aspects:~(a)~the maximal length of test traces cannot be reduced, and (b)~the
nondeterministic behaviour cannot be tested with smaller or fewer sets of
events, without losing the test suite's completeness property.

### 2018-10-07 paper/complexity.tex
Fixed a few typos
checked fixmes and new material

In this 
\fixme{In that section, $k$ is used for a different purpose. It is confusing.
I would use $p$ here.} 
\fxnote{jp: ok -- I also had to change this in Theorem~2, where $k$ was used as index from
0 to $pq-1$. I now use $j$ as index for $U_F(j)$, because $p$ is the number of states in $G(P)$.}
section, we
 will   calculate upper bounds for the total number of test executions to be performed
 when testing for failures refinement.   Theorem~\ref{th:failurestest} specifies
 that all tests $U_F(j),\ 0\le j < pq$ need to be executed, where $p$ denotes the
 number of nodes in the transition graph of the reference process $P$, and $q\ge p$ is an estimate for the number of nodes in the SUT's transition graph. Therefore, we will first calculate a bound for the number of test executions to be performed 
 for test $U_F(j)$ and then summarise  these bounds over all $j$ from $0$ to $pq-1$.

For the worst-case estimate, we
assume that $P$ never allows for early deadlock (so $\minhits(P/s)$ is never empty) 
and that the SUT $Q$ is a correct 
failures refinement. Then all test executions $(Q\parallel[\Sigma] U_F(j))$
stop after having run through a $Q$-trace of length $j+1$, because their is
no early termination due to entering branches (\ref{eq:ufa}), (\ref{eq:ufb}), or due to
an illegal deadlock of $Q$.
As can be seen from the specification of the test cases $U_F(j)$ (see
Section~\ref{sec:finitecompletefails}), the number of
executions ending in a $\epass$ event corresponds to the number $\ell$ of
traces $s$ of $P$ with length equal to  
$j$, multiplied by the number $h$ of minimal hitting sets in
$\minhits(P/s)$. For
the tests $U_T(j)$ verifying trace refinement (see
Section~\ref{sec:finitecomplete}), the number of executions equals $\ell$,
since there is no equivalent in $U_T(j)$ to checking different hitting sets
in the last step of a test execution.
\fixme{But doesn't this add to the number of executions?}
\fxnote{jp: should be clear now from the revised test}

@@

By Theorem~\ref{th:failurestest}, we need to execute the tests $U_F(j)$ for
$j = 0,\dots,(pq-1)$; this results in a worst-case bound defined below, where
we use the formula for the sum of the geometric progression.
%
\[
h\cdot \big( \frac{1-\card{\Sigma}^{pq}}{1-\card{\Sigma}} \big),\
 \text{or, asymptotically,}\  O(\frac{2^\card{\Sigma}}{\sqrt{2\pi \card{\Sigma}}}\cdot\card{\Sigma}^{(pq-1)}).
 \]
%
From 
\fixme{Theorem 4 is not
about $conf$. I don't think you have tests for $conf$.}
\fxnote{jp: I rephrased this -- hope that it's better understandable now.}
Theorem~\ref{th:upperboundh} we know that the worst-case bound for $h$
above cannot be further reduced, since the full collection of minimal
hitting sets needs to be checked at the end of each execution of $U_F(j)$.

@@

According 
\fixme{Please, see comments
above about this process.}\fxnote{jp: should be ok now after the above changes and clarifications.}
to Theorem~\ref{th:tracetest}, a complete test suite checking trace
refinement just contains the adaptive test case $U_T(pq-1)$. As derived for
$U_F(j)$ above, the number of executions performed by $(Q\parallel[\Sigma]
U_T(pq-1))$ is bounded by $\card{\Sigma}^{pq-1}$.

@@

\fixme{Confusing
use of $p$ and $q$ again.}
\fxnote{jp: fixed}

@@

\begin{example}\label{ex:pq}
Consider 
\fxnote{jp: I have updated the example with $P$ and $Q$ as used in the proof of the theorem. This should make thinks more readable.}
\fixme{The internal choice became external. To test for traces refinement, yo
put forward just one adaptive tests. So, there is more going on here.I think
it is because your tests are checking both traces and failures refinement.}
\fxnote{jp: no -- you are right: we want {\it internal} choice for $P$. With external
choice, this only shows that when testing for trace refinement, the length $pq$ is required.}
the CSP reference process $P$ and an erroneous implementation $Q$
specified as follows.

\begin{eqnarray*}
P & = &  P(0)
\\
P(k) & = & (k < p-1) \& \big( (a \then P(k)) \intchoice ( b \then P(k+1))\big)
\\ & & \extchoice
\\ & & (k = p-1) \& (a \then P(k))
\\
Q & = & Q(0)
\\
Q(k) & = & (k < q-1) \& \big( a \then Q(k+1)    \big) 
\\ & & \extchoice
\\ & & ( k = q-1)\& \big( a\then Q(0) \extchoice b\then Q(0)  \big)
\end{eqnarray*}

\medskip
The normalised transition graphs of $P$ and $Q$ are depicted in Fig.~\ref{fig:examplepq}
for the case $p=3,\ q=4$. Using FDR4, it can be shown for concrete values of
$p$ and $q$ that the  ``test passed conditions''
\[
(\epass\then\Stop) \lessdet_F (Q\parallel[\Sigma] U_F(j))\hide \Sigma
\]
and
\[
(\epass\then\Stop) \lessdet_T (Q\parallel[\Sigma] U_T(j))\hide \Sigma
\]
hold for $j = 0,\dots,pq-2$. This means, that none of the test cases $U_F(j)$
and $U_T(j)$ are capable of detecting failures and trace refinement violations, 
respectively, if they only check traces up to length $pq-1$ (recall that this
corresponds to $j\le pq-2$).

Process $Q$, however, neither conforms to $P$ in the failures refinement 
relation, nor in the trace
refinement relation. This can only be seen when executing the test $U_F(pq-1)$ and
$U_T(pq-1)$, respectively. These tests fail, so this shows that $P\not\lessdet_F Q$ and
$P\not\lessdet_T Q$ according to Theorem~\ref{th:failurestest} and Theorem~\ref{th:tracetest}. Moreover, this shows that
the maximal trace
length $pq$ to be investigated in the tests cannot be further reduced without losing
the completeness property of the test suites.
\xbox
\end{example}

\fixme{But, there is no argument that the proofs based on graph products
corresponds to this process. In particular the hiding does not seem to play a
role.}
\fxnote{jp: should be clear now, since we have explained in Section 3 that only 
the graph product $G(P)\times G(Q)$ is   needed in the proofs.}
%
Generalising Example~\ref{ex:pq}, it can be shown that for any pair
$2\le p,q \in\mathbb{N}$,
there exist reference processes $P$ with $p$ states
and implementation processes $Q$ with $q$ states, such that
a violation of the trace refinement property
can only be detected with a trace of length $pq$. This is proven in the following
theorem. In the proof, we use the processes $P$ and $Q$ introduced in Example~\ref{ex:pq}.

@@

To prove statement~5, we observe that the specification of $P$ implies (the 
expression $(s\cnt b)$ denotes the number of $b$-events occurring in trace $s$)
\[
\minaccs(P/s) = \left\{
\begin{array}{ll}
\{ \{a\}, \{b\} \} & \text{for all $s\in\trc(P)$ with $(s\cnt b) <p-1$.}
\\
\{ \{a\} \} & \text{for all $s\in\trc(P)$ with $(s\cnt b) = p-1$.}
\end{array}
\right.
\]
and
\[
\minaccs(Q/s) = \left\{
\begin{array}{ll}
\{ \{a\}  \} & \text{for all $s\in\trc(Q)$ with $\#s \neq 0\mod(q-1)$.}
\\
\{ \{a,b\} \} & \text{for all $s\in\trc(P)$ with $\#s = 0\mod(q-1)$.}
\end{array}
\right.
\]
As a  consequence, the minimal acceptance set $A_P = \{a\}$ which is
contained in every $\minaccs(P/s)$ fulfils $A_P \subseteq A_Q$ for
any $A_Q\in\minaccs(Q/s)$, when $s\in\trc(P)\cap \trc(Q)$. Now Lemma~\ref{lemma:tgtrcref}, 
(\ref{eq:failrefb}) can be applied to conclude that $Q\ conf\ P$.
\xbox
\end{proof}
%
Since Theorem~\ref{th:maxtracelen} just states that a violation of trace
refinement may remain undetected if only traces shorter than $pq$ are checked
during tests, it can also be applied to our trace refinement tests.
Therefore, test suites $\{ U_T(z) \}$ with $z<pq-1$ are not complete. It is
discussed in Section~\ref{sec:conc} how the number of test traces to be
executed by complete test suites for failures or trace refinement can still
be reduced {\it without} reducing the maximal length.

>>>>

In this section, we calculate upper bounds for the total number of test
executions to be performed when testing for failures refinement.
Theorem~\ref{th:failurestest} specifies that all tests $U_F(j),\ 0\le j < pq$
need to be executed, where $p$ denotes the number of nodes in the transition
graph of the reference process $P$, and $q\ge p$ is an estimate for the
number of nodes in the SUT's transition graph. Therefore, we will first
calculate a bound for the number of test executions to be performed for test
$U_F(j)$ and then summarise  these bounds over all $j$ from $0$ to $pq-1$.

For the worst-case estimate, we assume that $P$ never allows for early
deadlock (so $\minhits(P/s)$ is never empty) and that the SUT $Q$ is a
correct failures refinement. Therefore, all test executions
$(Q\parallel[\Sigma] U_F(j))$ stop after having run through a trace of $Q$ of
length $j+1$, because their is no early termination due to entering branches
(\ref{eq:ufa}), (\ref{eq:ufb}), or due to an illegal deadlock of $Q$. As can
be seen from the specification of the test cases $U_F(j)$ (see
Section~\ref{sec:finitecompletefails}), the number of executions ending in a
$\epass$ event corresponds to the number $\ell$ of traces $s$ of $P$ with
length equal to $j$, multiplied by the number $h$ of minimal hitting sets in
$\minhits(P/s)$. For the tests $U_T(j)$ verifying trace refinement (see
Section~\ref{sec:finitecomplete}), the number of executions equals $\ell$,
since there is no equivalent in $U_T(j)$ to checking different hitting sets
in the last step of a test execution. \fixme{But doesn't this add to the
number of executions?} \fxnote{jp: should be clear now from the revised test}

@@

By Theorem~\ref{th:failurestest}, we need to execute the tests $U_F(j)$ for
$j = 0,\dots,(pq-1)$; this results in a worst-case bound defined below, where
we use the formula for the sum of the geometric progression.
%
\[
h\cdot \big( \frac{1-\card{\Sigma}^{pq}}{1-\card{\Sigma}} \big),\
 \text{or, asymptotically,}\  O(\frac{2^\card{\Sigma}}{\sqrt{2\pi \card{\Sigma}}}\cdot\card{\Sigma}^{(pq-1)}).
 \]
%
From Theorem~\ref{th:upperboundh} we know that the worst-case bound for $h$
above cannot be further reduced, since the full collection of minimal hitting
sets needs to be checked at the end of each execution of $U_F(j)$.

@@

According to Theorem~\ref{th:tracetest}, a complete test suite checking trace
refinement just contains the adaptive test case $U_T(pq-1)$. As derived for
$U_F(j)$ above, the number of executions performed by $(Q\parallel[\Sigma]
U_T(pq-1))$ is bounded by $\card{\Sigma}^{pq-1}$.

@@

@@

\begin{example}\label{ex:pq}
Consider the CSP reference process $P$ and an erroneous implementation $Q$
specified as follows.
%
\begin{eqnarray*}
P & = &  P(0)
\\
P(k) & = & (k < p-1) \& \big( (a \then P(k)) \intchoice ( b \then P(k+1))\big)
\\ & & \extchoice
\\ & & (k = p-1) \& (a \then P(k))
\\
Q & = & Q(0)
\\
Q(k) & = & (k < q-1) \& \big( a \then Q(k+1)    \big)
\\ & & \extchoice
\\ & & ( k = q-1)\& \big( a\then Q(0) \extchoice b\then Q(0)  \big)
\end{eqnarray*}
%
The normalised transition graphs of $P$ and $Q$ are depicted in
Fig.~\ref{fig:examplepq} for the case $p=3,\ q=4$. Using FDR4, it can be
shown for concrete values of $p$ and $q$ that the  ``test passed conditions''
\[
(\epass\then\Stop) \lessdet_F (Q\parallel[\Sigma] U_F(j))\hide \Sigma
\]
and
\[
(\epass\then\Stop) \lessdet_T (Q\parallel[\Sigma] U_T(j))\hide \Sigma
\]
hold for $j = 0,\dots,pq-2$. This means that none of the test cases $U_F(j)$
and $U_T(j)$ are capable of detecting failures and trace refinement
violations, if they only check traces up to length $pq-1$
(recall that this corresponds to $j\le pq-2$).

Process $Q$, however, neither conforms to $P$ in the failures refinement
relation, nor in the trace
refinement relation. This can only be seen when executing the test $U_F(pq-1)$ and
$U_T(pq-1)$, respectively. These tests fail, so this shows that $P\not\lessdet_F Q$ and
$P\not\lessdet_T Q$ according to Theorem~\ref{th:failurestest} and Theorem~\ref{th:tracetest}. Moreover, this shows that
the maximal trace
length $pq$ to be investigated in the tests cannot be further reduced without losing
the completeness property of the test suites.
\xbox
\end{example}
%
Generalising Example~\ref{ex:pq}, it can be shown that for any pair
$2\le p,q \in\mathbb{N}$,
there exist reference processes $P$ with $p$ states
and implementation processes $Q$ with $q$ states, such that
a violation of the trace refinement property
can only be detected with a trace of length $pq$. This is proven in the following
theorem. In the proof, we use the processes $P$ and $Q$ introduced in Example~\ref{ex:pq}.

@@

To prove statement~5, we observe that the specification of $P$ implies (the
expression $(s\cnt b)$ denotes the number of $b$-events occurring in trace $s$)
\[
\minaccs(P/s) = \left\{
\begin{array}{ll}
\{ \{a\}, \{b\} \} & \text{for all $s\in\trc(P)$ with $(s\cnt b) <p-1$.}
\\
\{ \{a\} \} & \text{for all $s\in\trc(P)$ with $(s\cnt b) = p-1$.}
\end{array}
\right.
\]
and
\[
\minaccs(Q/s) = \left\{
\begin{array}{ll}
\{ \{a\}  \} & \text{for all $s\in\trc(Q)$ with $\#s \neq 0\mod(q-1)$.}
\\
\{ \{a,b\} \} & \text{for all $s\in\trc(P)$ with $\#s = 0\mod(q-1)$.}
\end{array}
\right.
\]
As a  consequence, the minimal acceptance set $A_P = \{a\}$ which is
contained in every $\minaccs(P/s)$ fulfils $A_P \subseteq A_Q$ for
any $A_Q\in\minaccs(Q/s)$, when $s\in\trc(P)\cap \trc(Q)$. Now Lemma~\ref{lemma:tgtrcref},
(\ref{eq:failrefb}) can be applied to conclude that $Q\ conf\ P$.
\xbox
\end{proof}
%
Since Theorem~\ref{th:maxtracelen} just states that a violation of trace
refinement may remain undetected if only traces shorter than $pq$ are checked
during tests, it can also be applied to our trace refinement tests.
Therefore, test suites $\{ U_T(j) \}$ with $j<pq-1$ are not complete. It is
discussed in Section~\ref{sec:conc} how the number of test traces to be
executed by complete test suites for failures or trace refinement can still
be reduced {\it without} reducing the maximal length.

### 2018-10-07 paper/conclusion.tex
checked fixmes and new material

In 
\fixme{From the proofs of the main theorems, and what you say below, I
thought you did have verdicts coming from deadlock, and you simply chose not
to mark it with an inc event.} 
contrast to that, our test cases specified in
Section~\ref{sec:finitecompletefails} and \ref{sec:finitecomplete} are
adaptive. This has the advantage that test executions $(Q\parallel[\Sigma]
U_F(j))$ for failures refinement never block before the final step specified
by branch (\ref{eq:ufd}), and so we do not need inconclusive test results.
It should be noted, however, that it is
necessary for our test verdicts to recognise also deadlocks in the final test
step and interpret them as FAIL, as described in
Section~\ref{sec:finitecompletefails}. In practice, this is realised by
adding a timeout event to the testing environment which indicates deadlock
situations. For real-time systems, this is an accepted technique, because the
SUT has to respond within a pre-defined latency interval, otherwise its
behaviour is considered to be blocked and regarded as a failure.
The tests executions $(Q\parallel[\Sigma] U_T(j))$ 
\fixme{Missing hiding.}
\fxnote{jp: Here, the hiding operator is not necessary: the possible test executions are really the traces of $Q$ parallel $U_T(j)$. The hiding is only needed to specify the verdict.}  
for trace refinement never
block at all before stopping after the verdict $\epass$ or $\efail$, unless the
SUT process $Q$ has an unexpected deadlock state. 
\fxfatal{And also, if the SUT blocks, the test should block,
no?}\fxnote{jp: you are right, I changed the text here and added explanations above after the verdict explanation
in formula (\ref{eq:passT}).}
Recall from the explanations given in Section~\ref{sec:finitecomplete} that the 
blocking situation also leads to passing the test, because 
$(\epass\then \Stop)\lessdet_T (Q\parallel[\Sigma] U_T(j))\hide\Sigma$ still holds
if $(Q\parallel[\Sigma] U_T(j))\hide\Sigma$ only produces the empty trace.

The adaptive behaviour of both our test case types $U_F(j)$ and $U_T(j)$,
however, induces the obligation to check that {\it all}
possible executions have been
performed before the test can be considered as passed. Typically, it is
therefore assumed that a \emph{complete testing
assumption}~\cite{hierons_testing_2004} \fixme{I think you need this for the
failures refinement tests as well.}\fxnote{jp: yes, I re-arranged the text accordingly.} holds, which means that every possible
behaviour of the SUT is performed after a finite number of test executions.
In practice, this is realised by executing each test several times, recording
the traces that have been performed, and using hardware or software coverage
analysers to determine whether all possible test execution behaviours of the
SUT have been observed. Therefore, adaptive test cases come at the price of
having to apply some grey-box testing techniques enabling us to decide
whether all SUT behaviours have been observed.

@@

\paragraph{Acknowledgements}
The authors would like to thank Bill Roscoe and Thomas Gibson-Robinson for
their advice on using the FDR4 model checker and for very helpful discussions
concerning the potential implications of this paper in the field of model
checking. We are
grateful to Li-Da Tong from National Sun Yat-sen University, Taiwan, for
suggesting the applicability of Sperner's Theorem in the context of the work
presented here. Moreover, we thank Adenilso Simao from the University of S{\~a}o Paulo
for several helpful suggestions. 
The work of Ana Cavalcanti is funded by the Royal Academy of
Engineering and UK EPSRC Grant EP/R025134/1.

>>>>

In \fixme{From the proofs of the main theorems, and what you say below, I
thought you did have verdicts coming from deadlock, and you simply chose not
to mark it with an inc event.} contrast to that, our test cases specified in
Section~\ref{sec:finitecompletefails} and \ref{sec:finitecomplete} are
adaptive. This has the advantage that test executions $(Q\parallel[\Sigma]
U_F(j))$ for failures refinement never block before the final step specified
by branch (\ref{eq:ufd}), and so we do not need inconclusive test results. It
should be noted, however, that it is necessary for our test verdicts to
recognise also deadlocks in the final test step and interpret them as FAIL,
as described in Section~\ref{sec:finitecompletefails}. In practice, this is
realised by adding a timeout event to the testing environment which indicates
deadlock situations. For real-time systems, this is an accepted technique,
because the SUT has to respond within a pre-defined latency interval,
otherwise its behaviour is considered to be blocked and regarded as a
failure. The tests executions $(Q\parallel[\Sigma] U_T(j))$ \fixme{Missing
hiding.} \fxnote{jp: Here, the hiding operator is not necessary: the possible
test executions are really the traces of $Q$ parallel $U_T(j)$. The hiding is
only needed to specify the verdict.}for trace refinement never block at all
before stopping after the verdict $\epass$ or $\efail$, unless the SUT
process $Q$ has an unexpected deadlock state. Recall from the explanations
given in Section~\ref{sec:finitecomplete} that the blocking situation also
leads to passing the test, because $(\epass\then \Stop)\lessdet_T
(Q\parallel[\Sigma] U_T(j))\hide\Sigma$ still holds if $(Q\parallel[\Sigma]
U_T(j))\hide\Sigma$ only produces the empty trace.

The adaptive behaviour of both our test case types $U_F(j)$ and $U_T(j)$,
however, induces the obligation to check that {\it all} possible executions
have been performed before the test can be considered as passed. Typically,
it is therefore assumed that a \emph{complete testing
assumption}~\cite{hierons_testing_2004} holds, which means that every
possible behaviour of the SUT is performed after a finite number of test
executions. In practice, this is realised by executing each test several
times, recording the traces that have been performed, and using hardware or
software coverage analysers to determine whether all possible test execution
behaviours of the SUT have been observed. Therefore, adaptive test cases come
at the price of having to apply some grey-box testing techniques enabling us
to decide whether all SUT behaviours have been observed.

@@

\paragraph{Acknowledgements}
The authors would like to thank Bill Roscoe and Thomas Gibson-Robinson for
their advice on using the FDR4 model checker and for very helpful discussions
concerning the potential implications of this paper in the field of model
checking. We are also grateful to Li-Da Tong from National Sun Yat-sen
University, Taiwan, for suggesting the applicability of Sperner's Theorem in
the context of the work presented here. Moreover, we thank Adenilso Simao for
several helpful suggestions. The work of Ana Cavalcanti is funded by the
Royal Academy of Engineering and UK EPSRC Grant EP/R025134/1.
